**OptiFit: a fast method for fitting amplicon sequences to existing OTUs**

### 2021-02-11 paper/manuscript.Rmd
paper update

**Running title:** Optimal Resolution

@@

- CRC is one of the most common cancers and a leading cause of cancer related death
- There is evidence that the microbiome has a role in CRC development/progression and could be useful for biomarker detection and diagnostics.
- Begum et al (mBio 2020) recently demonstrated effective application of machine learning (ML) to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml).
- A common question when applying ML methods to microbiome data is which method and taxonomic level is optimal.
- This analysis utilizes the reproducible framework developed by Begum et al to quantify which ML method and taxonomic level produce the best performing classifier for CRC data.

@@

## Results
- Across the five ML methods tested, model performance tends to increase with taxonomic level usually peaking around genus/otu level and dropping off slightly with ASVs.
- Random forest was consistently the top performer at most taxonomic levels.
  - RF might be more appropriate for microbiome analysis since its more suitable for zero inflated data (need to look into literature)
- Within the RF model, the highest AUCs were observed for family, genus, and otu level data with no significant difference between the three (Figure 1). ASV performed significantly lower than OTU but not family and genus levels.

>>>>

**Running title:** ASV level resolution is unnecessary for prediction of SRNs from microbiome data

@@

- CRC is one of the most common cancers and a leading cause of cancer related death in the United States.
- There is evidence that the microbiome has a role in CRC development/progression and could be useful for biomarker detection and diagnostics.
- Begum et al (mBio 2020) recently demonstrated effective application of machine learning (ML) to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml).
- A common question when applying ML methods to microbiome data is which method and taxonomic level is optimal.
- This analysis utilizes the reproducible framework developed by Begum et al to quantify which ML method and taxonomic level produce the best performing classifier for detecting SRNs from microbiome data.

@@

## Results
- Across the five ML methods tested, model performance tends to increase with taxonomic level usually peaking around genus/OTU level and dropping off slightly with ASVs.
- Random forest was consistently the top performer at most taxonomic levels.
  - RF might be more appropriate for microbiome analysis since its more suitable for zero inflated data (need to look into literature)
- Within the RF model, the highest AUCs were observed for family, genus, and OTU level data with no significant difference between the three (Figure 1). ASV performed significantly lower than OTU but not family and genus levels.

### 2021-02-11 paper/manuscript.tex
paper update

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{helvet}
\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\doublespacing
\usepackage[left]{lineno}
\linenumbers
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\author{}
\date{\vspace{-2.5em}}

@@

\hypertarget{optimal-resolution-for-microbiome-based-classification-of-colorectal-cancer}{%
\section{Optimal Resolution for Microbiome-Based Classification of
Colorectal
Cancer}\label{optimal-resolution-for-microbiome-based-classification-of-colorectal-cancer}}

@@

Courtney R. Armour, (ande? begum?), Patrick D. Schloss \({^\dagger}\)

@@

\({\dagger}\) To whom correspondence should be addressed:

@@

\hypertarget{abstract-max-250-words}{%
\subsection{Abstract (max 250 words)}\label{abstract-max-250-words}}

\hypertarget{importance-max-150-words}{%
\subsection{Importance (max 150 words)}\label{importance-max-150-words}}

@@

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

\begin{itemize}
\tightlist
\item
  CRC is one of the most common cancers and a leading cause of cancer
  related death
\item
  There is evidence that the microbiome has a role in CRC
  development/progression and could be useful for biomarker detection
  and diagnostics.
\item
  Begum et al (mBio 2020) recently demonstrated effective application of
  machine learning (ML) to microbiome based classification problems and
  developed a framework for applying ML practices in a more reproducible
  way (mikropml).
\item
  A common question when applying ML methods to microbiome data is which
  method and taxonomic level is optimal.
\item
  This analysis utilizes the reproducible framework developed by Begum
  et al to quantify which ML method and taxonomic level produce the best
  performing classifier for CRC data.
\end{itemize}

@@

\hypertarget{results}{%
\subsection{Results}\label{results}}

\begin{itemize}
\tightlist
\item
  Across the five ML methods tested, model performance tends to increase
  with taxonomic level usually peaking around genus/otu level and
  dropping off slightly with ASVs.
\item
  Random forest was consistently the top performer at most taxonomic
  levels.

  \begin{itemize}
  \tightlist
  \item
    RF might be more appropriate for microbiome analysis since its more
    suitable for zero inflated data (need to look into literature)
  \end{itemize}
\item
  Within the RF model, the highest AUCs were observed for family, genus,
  and otu level data with no significant difference between the three
  (Figure 1). ASV performed significantly lower than OTU but not family
  and genus levels.
\end{itemize}

@@

\hypertarget{conclusion}{%
\subsection{Conclusion}\label{conclusion}}

@@

\hypertarget{materials-and-methods}{%
\subsection{Materials and Methods}\label{materials-and-methods}}

\begin{itemize}
\tightlist
\item
  data from prior study \{baxter\}
\item
  mikropml package
\item
  pvalues as previously described \{begum\}
\end{itemize}

@@

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

@@

\hypertarget{figures}{%
\subsection{Figures}\label{figures}}

>>>>

\documentclass[]{article}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx}
% Redefine \includegraphics so that, unless explicit options are
% given, the image width will not exceed the width of the page.
% Images get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\ScaleIfNeeded{%
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother
\let\Oldincludegraphics\includegraphics
{%
 \catcode`\@=11\relax%
 \gdef\includegraphics{\@ifnextchar[{\Oldincludegraphics}{\Oldincludegraphics[width=\ScaleIfNeeded]}}%
}%
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

\author{}
\date{\vspace{-2.5em}}
\usepackage{helvet}
\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\doublespacing
\usepackage[left]{lineno}
\linenumbers

@@

\section{Optimal Resolution for Microbiome-Based Classification of
Colorectal
Cancer}\label{optimal-resolution-for-microbiome-based-classification-of-colorectal-cancer}

@@

Courtney R. Armour, (ande? begum?), Patrick D. Schloss ${^\dagger}$

@@

${\dagger}$ To whom correspondence should be addressed:

@@

\subsection{Abstract (max 250 words)}\label{abstract-max-250-words}

\subsection{Importance (max 150 words)}\label{importance-max-150-words}

@@

\subsection{Introduction}\label{introduction}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  CRC is one of the most common cancers and a leading cause of cancer
  related death
\item
  There is evidence that the microbiome has a role in CRC
  development/progression and could be useful for biomarker detection
  and diagnostics.
\item
  Begum et al (mBio 2020) recently demonstrated effective application of
  machine learning (ML) to microbiome based classification problems and
  developed a framework for applying ML practices in a more reproducible
  way (mikropml).
\item
  A common question when applying ML methods to microbiome data is which
  method and taxonomic level is optimal.
\item
  This analysis utilizes the reproducible framework developed by Begum
  et al to quantify which ML method and taxonomic level produce the best
  performing classifier for detecting SRNs from microbiome data.
\end{itemize}

@@

\subsection{Results}\label{results}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Across the five ML methods tested, model performance tends to increase
  with taxonomic level usually peaking around genus/otu level and
  dropping off slightly with ASVs.
\item
  Random forest was consistently the top performer at most taxonomic
  levels.
\item
  RF might be more appropriate for microbiome analysis since its more
  suitable for zero inflated data (need to look into literature)
\item
  Within the RF model, the highest AUCs were observed for family, genus,
  and otu level data with no significant difference between the three
  (Figure 1). ASV performed significantly lower than OTU but not family
  and genus levels.
\end{itemize}

@@

\subsection{Conclusion}\label{conclusion}

@@

\subsection{Materials and Methods}\label{materials-and-methods}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  data from prior study \{baxter\}
\item
  mikropml package
\item
  preprocessing
\item
  hp tuning
\item
  n seeds
\item
  pvalues as previously described \{begum\}
\end{itemize}

@@

\subsection{Acknowledgements}\label{acknowledgements}

@@

\subsection{Figures}\label{figures}
